---
title: "ReactomeGraph4R: an R Interface for the Reactome Graph Database"
author: Chi-Lam Poon
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float:
      collapsed: false
    highlight: pygments
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview
Pathways, reactions, and biological entities in Reactome knowledge are systematically represented as an ordered network of molecular reactions. Graph database technology is an effective tool for modeling highly connected data, hence Reactome's relational database is imported in [Neo4j](https://neo4j.com/) to create one large interconnected graph. Instances are represented as _nodes_ and relationships between nodes as _edges_.

The [`ReactomeGraph4R`](https://github.com/reactome/ReactomeGraph4R) package is an R interface for retrieving **data with network structure** from the [Reactome Graph Database](https://reactome.org/dev/graph-database). There is another R package, [`ReactomeContentService4R`](https://github.com/reactome/ReactomeContentService4R), for querying **specific bits of information** from the Reactome Database through the RESTful API in the Content Service.

`ReactomeGraph4R` is built on the Neo4j driver [`neo4r`](https://github.com/neo4j-rstats/neo4r), thus returned data are mainly same as those called by `neo4r` but with a little modifications, and are in these two formats:

- "row": a list of results in dataframes
- "graph": a graph object with `nodes` and `relationships` information that can be used for visualization

This package will allow you to interact with the data in Reactome's graph database in R, with the aim of minimizing the number of Neo4j [`Cypher`](https://neo4j.com/developer/cypher/) queries that the user will need to perform. For example, if you wanted to retrieve any Reactome information associated with the hypothetical identifier '123456789', you can use  `matchObject(id="123456789")`, which would be equivalent to using the Cypher query `MATCH (rgp:ReferenceGeneProduct) WHERE rgp.identifier = "123456789" RETURN rgp` on the Reactome graph database. 

Aside from performing basic Cypher queries and formatting the results as R objects, the package also contains functionality that can not be easily performed using Cypher. This includes finding hierarchical data of an instance (for example what Reactions and Pathways a Protein is found in), getting the entire Reaction sequence/context using preceding/following relationships, what role a PhysicalEntity plays in each of its associated Reactions (catalyst, regulator, input, etc.), searching for research papers that are cited in Reactome's curations, and even displaying network data. Please read on to see detailed instructions for the `ReactomeGraph4R` package - it is a flexible package with plenty of useful functionality for the prospective R-Reactome user!

## Setups

Follow this [instruction](https://github.com/reactome/ReactomeGraph4R/blob/develop/README.md) to download and setup the Reactome Graph Database.

Then install `ReactomeGraph4R` package:
```
devtools::install_github("reactome/ReactomeGraph4R")
```

There are two questions needed to be answered for Neo4j server connection when loading the package. You can change the url port if it's not `7474`. And if the Neo4j authentication is required, the username and password are same as the ones to log in your local Neo4j database.

```{r setup, eval=FALSE}
library(ReactomeGraph4R)
login()
```

```
## Is the url 'http://localhost:7474'? (Yes/no/cancel) 
## Does Neo4J require authentication? (yes/No/cancel)
```

```
## Successfully connected to the local Reactome Graph Database v76!
```

## Basic query

The basic function `matchObject` allows you to fetch Reactome objects using:

- `id`: Reactome or non-Reactome identifier (e.g. UniProt id)
- `displayName`: display name of an object 
- `schemaClass`: schema class 
- `property`: attributes of Reactome objects
- `relationship`: relationship between two nodes 

Moreover, you could specify the argument `returnedAttributes` for retrieving only a few attributes of the targeted object; `species` for specific species; and `limit` for the number of returned objects. 
Note that this function only returns "row" data.

### Fetch by id

The "id" input can be either non-Reactome or Reactome identifiers. If you use a non-Reactome id, remember that you _must_ also specify `databaseName` since the default one is "Reactome". For example, to get the Reactome instance associated with a circadian rhythmic gene _PER2_:

```{r basic-id-1, eval=FALSE}
# Retrieve the object of PER2 gene
# NOTE: if you're unsure which database to specify, you have to leave it as NULL
matchObject(id = "PER2", databaseName = NULL)
```

```
## $databaseObject
##            schemaClass identifier   databaseName              displayName
## 1 ReferenceDNASequence       PER2 COSMIC (genes) COSMIC (genes):PER2 PER2
##       dbId geneName                                                     url
## 1 11509503     PER2 http://cancer.sanger.ac.uk/cosmic/gene/overview?ln=PER2
```

Now we know that the database name should be "COSMIC (genes)"! We can also try with a Reactome id "R-HSA-400219":

```{r basic-id-2, eval=FALSE}
matchObject(id = 'R-HSA-400219')
```

```
## $databaseObject
##   schemaClass  speciesName     oldStId isInDisease releaseDate
## 1    Reaction Homo sapiens REACT_25088       FALSE  2010-09-21
##                                    displayName    stIdVersion   dbId
## 1 Beta-TrCP1 binds phosphorylated PER proteins R-HSA-400219.1 400219
##                                           name         stId category isInferred
## 1 Beta-TrCP1 binds phosphorylated PER proteins R-HSA-400219  binding       TRUE
```

For multiple ids, say you want to get more information for your significantly enriched pathways, you can use function `multiObjects`. The `speedUp` option can determine to use the [doParallel](https://cran.r-project.org/web/packages/doParallel/index.html) method or not, details see `?multiObjects`.

```{r multi-ids, eval=FALSE}
# retrieve multiple objects
ids <- c('R-HSA-74158', 'R-HSA-1566977', 'R-HSA-3000157', 'R-HSA-3000178', 'R-HSA-216083')
multiObjects(ids)
```

### Fetch by name

Instances can also be fetched by their "displayNames". Do note that **spaces** and **symbols** within the name are required. Here we focus on the complex SUMO1:TOP1 in nucleoplasm "SUMO1:TOP1 [nucleoplasm]" in _C. elegans_:

```{r basic-name, eval=FALSE}
matchObject(displayName = "SUMO1:TOP1 [nucleoplasm]", species = "C. elegans")
```

```
## $databaseObject
##   schemaClass            speciesName isInDisease              displayName
## 1     Complex Caenorhabditis elegans       FALSE SUMO1:TOP1 [nucleoplasm]
##       stIdVersion     dbId       name          stId
## 1 R-CEL-4641301.1 10549504 SUMO1:TOP1 R-CEL-4641301
```

### Fetch by class

When retrieving instances belonging to one schema class, it's better specify the argument `limit` as well for restricting the number of returned instances. For all available schema classes see Reactome [Data Schema](https://reactome.org/content/schema). For instance, to get 5 "EntitySets" in human and then return their display names and stId only:

```{r basic-class, eval=FALSE}
# Get 5 instance in Class EntitySet and return displayName & stId
entity.set <- matchObject(schemaClass = "EntitySet", species = "human",
                          returnedAttributes = c("displayName", "stId"), limit = 5)
entity.set[["databaseObject"]] # show as dataframe
```

### Fetch by property

By specifying the `property`, nodes with the given property (or properties), which are actually attributes/slots of Reactome instances, could be returned. Let's try to get instances that are chimeric and are in disease.

```{r basic-property, eval=FALSE}
# Get instances with conditions of properties that are stored in a list
matchObject(property = list(isChimeric = TRUE, isInDisease = TRUE), limit = 10)[["databaseObject"]]
```

### Fetch by relationship

The actual Cypher query for this command is `MATCH (n1)-[r:relationship]->(n2) RETURN n1,n2`, therefore the `n1` and `n2` dataframes in the returned list have the same number of rows, and every two rows with the same index are connected with the given relationship.

```{r basic-rel, eval=FALSE}
# Get nodes connected with 'hasComponent'
matchObject(relationship = "hasComponent", limit = 3)
```


## MATCHing

These following functions in the __MATCH__ family provide several commonly used cases that you might be interested in for Reactome data querying.


### Hierarchy data 

Reactome data are organized in a hierarchical way: `Pathway --> Reaction --> PhysicalEntity`, or sometimes it might be `Pathway --> Reaction --> PhysicalEntity --> ReferenceEntity` where the PhysicalEntity has links to external database information via the ReferenceEntity. You could retrieve the hierarchical data of a given __Event__ (Pathway or Reaction) or __Entity__ (PhysicalEntity or ReferenceEntity) using `matchHierarchy`. In this example, we'll take a look at a RNA sequence (PhysicalEntity) "POU5F1 mRNA [cytosol]" with stable identifier "R-HSA-500358":

```{r hierarchy, eval=FALSE}
# Get hierarchy data of R-HSA-500358
pou5f1.hierarchy <- matchHierarchy(id = "R-HSA-500358", type = "row")
str(pou5f1.hierarchy, max.level = 1)
```

The RNA sequence we specified is in the `physicalEntity` dataframe of the result list. It's directly connected with those Events in the `event` dataframe, which are then connected with Events in the `upperevent`. Relationships between all these objects are in `relationship` dataframe:

```{r hierarchy-2, eval=FALSE}
# select essential columns to show
pou5f1.hierarchy$relationships[,c(2,4,5,7,8)]
```


### Reactions in associated Pathway

This method can find all __ReactionLikeEvents__ (RLEs) connected with a given Pathway by the relationship "hasEvent". Additionally, the input can be a RLE, the result would be Pathway(s) linked via "hasEvent" together with other RLEs linked with the Pathways(s). Here we focus on a RLE "OAS1 oligomerizes" with identifier "R-HSA-8983688".

```{r r-in-p, eval=FALSE}
# Find Reactions connected with R-HSA-8983688
rle <- matchReactionsInPathway(event.id = "R-HSA-8983688", type = "row")
str(rle, max.level = 1)

# The one in reactionLikeEvent is what we search for
rle$reactionLikeEvent

# Take a look at the connected Pathway
rle$pathway
```

`otherReactionLikeEvent` are RLEs other than "OAS1 oligomerizes" connected with Pathway "OAS antiviral response".

```{r r-in-p-2, eval=FALSE}
# Show displayNames of other RLEs
rle$otherReactionLikeEvent[["displayName"]]
```

The contect of these Events can actually be visualized in R using the `exportImage` function from the `ReactomeContentService4R` package! And it looks the same as that in [Pathway Browser](https://reactome.org/PathwayBrowser/#/R-HSA-8983711&PATH=R-HSA-168256,R-HSA-1280215,R-HSA-913531,R-HSA-1169410). 
To get the pathway diagram of Pathway "OAS antiviral response" (stId: R-HSA-8983711) that we just retrieved, and highlight the RLE (stId: R-HSA-8983688) that we specified:

```{r export-img, eval=FALSE}
library(ReactomeContentService4R)

# Export pathway diagram of "OAS antiviral response"
exportImage(id = "R-HSA-8983711", output = "diagram", 
            sel = "R-HSA-8983688", format = "png", quality = 8)
```


### Preceding/following Events 

With the diagram shown above, we can see that the Reaction highlighted in blue is in the middle of a Reaction cascade, with other RLEs immediately preceding and following it. In order to know what these preceding and following Reactions are, we can use function `matchPrecedingAndFollowingEvents` to find RLEs linked via "precedingEvent". The argument `depth` is used to describe the "variable length relationships", the default value is 1 (i.e. immediately connected); or you can set `all.depth = TRUE` for retrieving the whole context. Details see `?matchPrecedingAndFollowingEvents`.

```{r p-f-event-1, eval=FALSE}
# Retrieve RLE context with depth = 2
rle.context <- matchPrecedingAndFollowingEvents(event.id = "R-HSA-8983688", depth = 2, type = "row")
str(rle.context, max.level = 1)
```


### Referrals

Usually we query data in a way like parent to child `(parent) --> (child)`, where we provide information about the parent. But with the Graph Database, we are able to search in a reverse direction that is child to parent `(parent) <-- (child)` with child's information only. This "child-to-parent" relationship is called __Referral__. You could carry out the referral fetching by `matchReferrals` that supports Classes "Event", "PhysicalEntity", "Regulation", "CatalystActivity", "ReferenceEntity", "Interaction", "AbstractModifiedResidue". Depth related arguments could also be specified here. More details sees `?matchReferrals`.

We would look at a Regulation "Negative gene expression regulation by 'EGR2 [nucleoplasm]" with dbId "6810147":

```{r referrals, eval=FALSE}
# Find referrals of the given Regulation
matchReferrals(id = 6810147, type = "row")
```
The dbId of _endNode_ (`endNode.dbId` in `$relationships`) is exactly the dbId we just specified.


### Interactors

Interactions of a PhysicalEntity (PE) could be retrieved by `matchInteractors`. This method begins with finding the ReferenceEntity matched with the PE, then get the Interactions having "interactor" relationship with the ReferenceEntity. For example, to get interactions of "FANCM [nucleoplasm]" with stable id "R-HSA-419535":

```{r interactor, eval=FALSE}
# Retrieve interaction data of the given PE
interactors <- matchInteractors(pe.id = "R-HSA-419535")
str(interactors, max.level = 1)
interactors$interaction
```


### PhysicalEntity roles

The roles of PhysicalEntities include "input", "output", "regulator", "catalyst", which are represented as relationships "input" ,"output", "regulatedBy", "catalystActivity" respectively. Therefore, we could retrieve instances that are possibly connected with the given PhysicalEntity via these relationships, and see the exact role(s) from the existing relationships. 
We'll take a look at a Polymer "HSBP1 oligomer [cytosol]" and input it into `matchPEroles`. Either `id` or `displayName` could be specified.

```{r, eval=FALSE}
# Find possible roles of the given PE
roles <- matchPEroles(pe.displayName = "HSBP1 oligomer [cytosol]")
str(roles, max.level = 1)

# get the roles (relationships type)
unique(roles$relationships$type)
```


### Diseases

Diseases related to a PhysicalEntity or an Event could be found using function `matchDisease`. In reverse, you can also get PhysicalEntities/Events associated with a Disease.

```{r disease, eval=FALSE}
# Fetch Reactome instances associated with 'neuropathy' in human
matchDiseases(displayName = "neuropathy", species = "human", type = "row")
```


### Papers

Given the PubMed id or the title for a paper, Reactome instances related to this paper could be found by `matchPaperObjects`. The DatabaseObjects are connected with the LiteratureReference (i.e. paper) via "literatureReference" relationship. Let's try with a paper "Aggresomes: a cellular response to misfolded proteins".

```{r paper, eval=FALSE}
# fetch objects by paper title
matchPaperObjects(displayName = "Aggresomes: a cellular response to misfolded proteins", type = "row")
```


## Network graph

The ability to view network graphs is definitely a big advantage of a graph database. Fortunately, R has developed into a powerful tool for network analysis. There are a number of R packages targeted network analysis and visualization, therefore we are able to get a graph just like the one in the Neo4j server, and even to set more visualization options! 

Don't forget that results can also be returned in the "graph" format, which are used to create the network visualization in R! This comprehensive tutorial - [_Network visualization with R_](https://kateto.net/network-visualization) (Ognyanova, K., 2019) - walks through each step on the creation of network graphs in R. 

Here we will show a couple of examples to generate an interactive network graph after retrieving the specific Reactome graph data. Let's say we want to visualize the hierarchical data of a ReferenceEntity "UniProt:P33992 MCM5".

First install and load the following packages.

### Package installation

```{r vis-loading, eval=FALSE}
# install packages
list.pkg <- c("stringr", "visNetwork", "networkD3", "wesanderson")
new.pkg <- list.pkg[!(list.pkg %in% installed.packages()[ ,"Package"])]
if (length(new.pkg)) {
  install.packages(new.pkg, repos = "https://cloud.r-project.org/")
}
# load
invisible(suppressPackageStartupMessages(lapply(list.pkg, library, character.only = TRUE)))
```


### visNetwork

We will try the [`visNetwork`](https://datastorm-open.github.io/visNetwork/) package which visualizes networks using [`vis.js`](http://visjs.org/) javascript library.

```{r vis-1, eval=FALSE}
# Get graph output data
graph <- matchHierarchy(displayName = "UniProt:P33992 MCM5", databaseName = "UniProt", type = "graph")
relationships <- graph[["relationships"]]
nodes <- graph[["nodes"]]
nodes <- unnestListCol(df = nodes, column = "properties") # unnest the 'properties' column of lists

head(nodes); head(relationships)


# Transform into visNetwork format for nodes & edges
vis.nodes <- data.frame(id = nodes$id, 
                        label = str_trunc(nodes$displayName, 20), # truncate the long names
                        group = nodes$schemaClass, 
                        title = paste0("<p><b>", nodes$schemaClass, "</b><br>", 
                                "dbId: ", nodes$dbId, "<br>", nodes$displayName, "</p>"))

vis.edges <- data.frame(from = relationships$startNode,
                        to = relationships$endNode,
                        label = relationships$type,
                        font.size = 16,
                        font.color = 'steelblue')

head(vis.nodes); head(vis.edges)
```

We are going to change the visual parameters of nodes and edges by adding them as columns in the dataframes. More customizations see the `visNetwork` [documentation](http://datastorm-open.github.io/visNetwork/) or `?vignette("Introduction-to-visNetwork")`.

```{r vis-2, eval=FALSE}
# nodes parameters
## get palette colors with package 'wesanderson'
node.colors <- as.character(wes_palette(n = length(unique(vis.nodes$group)), name = "Darjeeling2"))
names(node.colors) <- levels(factor(vis.nodes$group))
## NOTE: don't use `str_replace_all` here since 'TopLevelPathway' & 'Pathway' share the string 'Pathway' 
vis.nodes$color.background <- node.colors[as.numeric(factor(vis.nodes$group))] # node color
vis.nodes$color.border <- "lightgray"
## highlight the instance we specified
vis.nodes$color.border[vis.nodes$label == "UniProt:P33992 MCM5"] <- "pink" 
vis.nodes$color.highlight.border <- "darkred"
vis.nodes$borderWidth <- 2 # Node border width

# edges parameters
vis.edges$width <- 1.2 # line width
edges.colors <- as.character(wes_palette(n = length(unique(vis.edges$label)), name = "FantasticFox1"))
names(edges.colors) <- unique(vis.edges$label)
vis.edges$color <- str_replace_all(vis.edges$label, edges.colors) # line color  
vis.edges$arrows <- "to" # arrows: 'from', 'to', or 'middle'
vis.edges$smooth <- TRUE    # should the edges be curved?

# height & width of the plot can be set here
visnet <- visNetwork(vis.nodes, vis.edges, main = "The hierarchy of protein MCM5", 
                     height = "500px", width = "100%")
visnet
```

Add a drop-down menu:

```{r vis-drop-down, eval=FALSE}
# Rename column name 'group' to 'Class' for displaying in the window
names(visnet[["x"]][["nodes"]]) <- gsub("group", "Class", names(visnet[["x"]][["nodes"]]))
visOptions(visnet, highlightNearest = TRUE, selectedBy = "Class")
```


### networkD3

We can also take a look at another package [`networkD3`](http://christophergandrud.github.io/networkD3/), which generates network graphs using [D3](http://d3js.org/) javascript library.

```{r d3-1, eval=FALSE}
# the node ids MUST be numeric, and start from 0
nodes.idx <- as.character(as.numeric(factor(nodes$id)) - 1)
names(nodes.idx) <- nodes$id

# transform into networkD3 format
d3.edges <- data.frame(source = as.numeric(str_replace_all(relationships$startNode, nodes.idx)),
                       target = as.numeric(str_replace_all(relationships$endNode, nodes.idx)),
                       label = relationships$type)
d3.edges <- d3.edges[order(d3.edges$source), ]

d3.nodes <- cbind(idx=as.numeric(nodes.idx), nodes)
d3.nodes <- d3.nodes[order(d3.nodes$idx), ] # the order MUST be consistent with the 'source'

forceNetwork(Links = d3.edges, Nodes = d3.nodes, Source="source", Target="target",
             NodeID = "displayName", Group = "schemaClass", Value = "label", 
             linkColour = "#afafaf", fontSize = 12, zoom = TRUE, legend = TRUE,
             Nodesize = 15, opacity = 0.9, charge = -50)
```

To modify the forceNetwork graph, one can execute custom javascript code with the [`htmlwidgets`](https://www.htmlwidgets.org/) R package, but it won't be discussed here.


## SessionInfo

```{r secinfo}
sessionInfo()
```
